Promotors' reports on the first draft of the thesis document.

### 2020-08-21 ChapterStatisticalAnalysis/ChapterStatisticalAnalysis.tex
Second draft before public defense

\begin{quote}
%\vspace{-0.5cm}
\emph{The results of this chapter were published in Quintana Carapia, G., Markovsky, I. Pintelon, R., Csurcsia, P.Z., and Verbeke, D., "Bias and covariance of the least-squares estimate in a structured errors-in-variables problem", Computational Statistics & Data Analysis journal, Vol. 144, 2020, ISSN 0167-9473, doi:10.1016/j.csda.2019.106893. \nocite{QuintanaCSDA} }\vfill{}
\end{quote}

@@

The linear estimation problems with perturbation in both the regression matrix and the regressor are errors-in-variables (EIV) problems.
It is required to calculate the statistical moments of the estimate to evaluate the properties of the solution of an EIV problem.
The calculation of the estimate statistical moments demands a more considerable effort when the regression matrix of an EIV problem is structured, and when the regression matrix and the regressor are correlated.

Structured total least-squares and structured total maximum likelihood are two examples of methods that solve structured EIV problems, but that require large computational resources.
For online applications, where a real-time solution of the structured EIV problem is imperative, the recursive least-squares (RLS) method provides a suboptimal but computationally cheaper online solution.

Statistical analysis is indispensable to obtain the statistical moments of the RLS solution for a structured and correlated EIV problem, and to assess the estimation uncertainty.
This chapter describes the methodology and the results of the statistical analysis.
The mathematical expectation of the RLS estimate, approximated by a second-order Taylor series expansion, allows a quantification of the first and second-order moments of the LS estimate.
The statistical analysis results provide insight into the impact that the structure of the regression matrix and its correlation with the regressor has on the uncertainty of the RLS estimate.

\begin{comment}
Errors-in-variables (EIV) are linear estimation problems in which the regression matrix and the regressor are perturbed \citep{VanHuffel91Book}, \citep{Markovsky07overview}.
In structured EIV problems, the regression matrix has a given structure that depends on the problem formulation.
Hankel, Toeplitz, or other application-specific matrices appear in problems of metrology \citep{Markovsky15cep}, system identification \citep{Soderstrom07}, image restoration \citep{Feiz17}, nuclear magnetic resonance spectroscopy \citep{Cai16}, direction-of-arrival estimation \citep{Pan18}, and time-of-arrival estimation \citep{Jia18}.

In metrology, the direct estimation of the input from the sensor transient response is formulated as a structured EIV problem.
The only observed signal is the sensor output.
To estimate a step input, the regression matrix, and the regressor are built from the step response observations \citep{Markovsky15cep}, and
the structure in the regression matrix is block-Hankel.
This data-driven estimation methodology reduces the estimation time of the classical two-stage approach where a sensor model is first identified, and later the input is estimated using the sensor model \citep{Azam15, Niedzwiecki16a}.
%The problem is EIV because the perturbation of the sensor response gets into the regression matrix.

Total least-squares (TLS) is the typical estimator for unstructured EIV problems and is consistent when the perturbations have zero mean, and the covariance is a given positive definite matrix.
When the perturbations are i.i.d. normally distributed, the solution of the TLS is equivalent to that of the maximum likelihood estimator (ML) \citep{Markovsky07overview}. 
For structured EIV problems, the TLS estimator does not give general results since each specific structure requires a particular treatment \citep{VanHuffel07TLSeditorial}, and the ML estimator leads to non-convex optimization problems where finding the global optimum is not guaranteed \citep{Rhode14recursive}.
Moreover, the computational complexity of TLS and ML inhibits real-time implementation to solve structured EIV problems.

The least-squares (LS) estimator is a suboptimal but simple solution to structured EIV problems that admits a recursive form for easy real-time implementations.
Some of the reported works that propose LS estimators for structured EIV problems include 
the design of a fast algorithm for matrices with small displacement rank \citep{Mastronardi07fast},
the study of the estimator consistency \citep{Palanthandalam10parameter},
the determination of the bias, and the mean squared error of the parameter estimates in the identification of AR models \citep{Kiviet12high} \citep{Kiviet14improved}, and
a discussion of the causes of bias and inconsistency in homogeneous estimators \citep{Yeredor04homogeneous}.

 In measurement applications, it is highly relevant to assess the uncertainty of the input estimate.
The uncertainty of the reported LS estimators for structured EIV problems has not been addressed, and then, remains unknown.
The estimator uncertainty is expressed using the estimation bias and covariance \citep{Pintelon12Book}.
To know the LS estimator uncertainty, we quantified the bias and the covariance of the LS solution of EIV problems in the unstructured and structured cases. 
This extends the perturbation analysis of the LS estimator of unstructured and uncorrelated problems that was investigated in \citep{Stewart90SPT} and in \citep{Vaccaro94}.
This paper presents a study of the statistics of the LS estimate of unstructured and structured EIV problems. 
 We provide a discussion of the unstructured case as a reference, to get an insight of the impact that the structure of the regression matrix and the correlation between the regression matrix and the regressor perturbations have on the uncertainty of the LS estimate.
The structured case is motivated by the metrology estimation problem \citep{Markovsky15cep}, where the EIV problem has a Hankel structure, and the perturbations are correlated.
The study of the metrology input estimate illustrates a methodology to conduct statistical analysis for any structured EIV problem.
The mathematical expectation of the second-order Taylor series expansion of the LS estimate boils down to expressions that quantify the first and second-order moments of the LS estimate.
Via Monte Carlo simulations, we validated the accuracy of the bias, and the covariance approximations.
The derived approximations predict the LS estimation bias, and the covariance, for given sample size and perturbation level.
The predicted variance gives the uncertainty of the LS estimate.
We observed that, for the step input estimation problem, the mean squared error of the LS estimate is near to the minimum variance limit given by the Cram\'er-Rao lower bound of the structured EIV problem.
 By following this methodology, the bias and variance of the solutions of EIV problems with other structures is determined, and therefore, the uncertainty of the estimate.
\end{comment}

@@

For an overdetermined system of linear equations, constructed by the data-driven step input estimation method (\ref{eqn:min_seiv}), the least-squares (LS) solution is given by
\begin{equation} \widehat{\bm{\theta}} = \widetilde{\mathbf{K}}^\dagger \widetilde{\mathbf{y}} = ( \widetilde{\mathbf{K}}^\top \widetilde{\mathbf{K}} )^{-1} \widetilde{\mathbf{K}}^\top \widetilde{\mathbf{y}} , \label{eqn:xhat} \end{equation}
where $\widetilde{\mathbf{K}}^\dagger$ is the pseudo-inverse of $\widetilde{\mathbf{K}}$.
The objective of the statistical analysis is to obtain the bias and the covariance of the solution $\widehat{\bm{\theta}}$ to the structured and correlated EIV problem (\ref{eqn:ddsiemnd}).
The bias and the covariance are computed using the mathematical expectation operator $\mathbb{E}\{\cdot\}$.
First we substitute the expressions (\ref{eqn:y0plusnoise}) and (\ref{eqn:K0plusnoise}) in Equation (\ref{eqn:xhat}) to obtain
\begin{equation*} \widehat{\bm{\theta}} = \left( (\mathbf{K}+\mathbf{E})^\top (\mathbf{K}+\mathbf{E})  \right)^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}), \end{equation*} 
which is equivalent to
\begin{equation} \widehat{\bm{\theta}} = \left( \mathbf{I} + \mathbf{M} \right)^{-1} \mathbf{Q}^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}), \label{eqn:1steq} \end{equation} 
where 
\begin{equation} \mathbf{Q} = \mathbf{K}^\top \mathbf{K}, \quad \text{and} \quad \mathbf{M} = \mathbf{Q}^{-1} ( \mathbf{K}^\top \mathbf{E} + \mathbf{E}^\top \mathbf{K} + \mathbf{E}^\top \mathbf{E} ). \end{equation} 
Applying a second-order Taylor expansion of the inverse matrix
\begin{equation} (\mathbf{I} + \mathbf{M})^{-1} \approx \mathbf{I} - \mathbf{M} + \mathbf{M}^2, \label{eqn:TseriesExp} \end{equation} 
that is valid when the SNR is sufficiently high, and therefore $\mathbf{E}$ and $\mathbf{M}$ are small, satisfying the constraint on the spectral radius $\| \mathbf{M} \| < 1$. 
The neglected term in the Taylor series expansion is of the order $O(\Vert M \Vert^3)$.
We can express the estimate as
\begin{equation} \widehat{\bm{\theta}} \approx \left( \mathbf{I} - \mathbf{M} + \mathbf{M}^2 \right) \mathbf{Q}^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}). \label{eqn:xhatexp} \end{equation}

Now that the perturbation variables $\bm{\epsilon}$ and $\mathbf{E}$ are no longer inside an inverse matrix, we can compute the mathematical expectation of expressions derived from the Taylor series approximation (\ref{eqn:xhatexp}) of $\widehat{\bm{\theta}}$. 
The bias of the estimate $\widehat{\bm{\theta}}$ is obtained from
\begin{equation} \mathbf{b} \left(\widehat{\bm{\theta}} \right) = \bm{\mu}\left(\widehat{\bm{\theta}} \right) - \bm{\theta}, \label{eqn:biasdef} \end{equation}
where $\bm{\mu} \left(\widehat{\bm{\theta}} \right) = \mathbb{E} \left\{ \widehat{\bm{\theta}} \right\}$, and $\bm{\theta} = \mathbf{K}^\dagger \mathbf{y} = \mathbf{Q}^{-1} \mathbf{K}^\top \mathbf{y}$ is the true value.
The covariance of the estimate $\widehat{\bm{\theta}}$ is obtained from
\begin{equation} \begin{aligned} \mathbf{C} \left( \widehat{\bm{\theta}} \right) & = \mathbb{E} \left\{ \left( \widehat{\bm{\theta}} - \bm{\mu} \right) \left( \widehat{\bm{\theta}} - \bm{\mu} \right)^\top \right\} . \end{aligned} \label{eqn:covdef} \end{equation}

The terms derived from (\ref{eqn:xhatexp}) that do not contribute to the bias and to the covariance are filtered out by the mathematical expectation operator considering the following general rules that are valid regardless of the existence of structure in the regression problem: 
\begin{itemize}
	\item the expected values $\mathbb{E} \{ \mathbf{E} \} = \mathbf{0}$, and $\mathbb{E} \{ \bm{\epsilon} \} = \mathbf{0}$, since $\mathbf{E}$ and $\bm{\epsilon}$ are zero-mean random variables, and
	\item the expected values of odd order moments, such as $\mathbb{E} \{ \mathbf{E}^\top \mathbf{E} \mathbf{E}^\top \}$, are zero.
\end{itemize}
Moreover, the second-order approximation disregards moments of order four and higher.

@@

First, the statistics of the least-squares estimator (\ref{eqn:xhat}) of an unstructured EIV problem is discussed, 
provided that the perturbations of the regression matrix and the regressor are i.i.d. normally distributed with zero mean, and variances $\sigma_{\mathbf{E}}^2$ and $\sigma_{\bm{\epsilon}}^2$, respectively.
Therefore, the terms in the Taylor series expansion (\ref{eqn:xhatexp}) that contain products of $\mathbf{E}$ and $\bm{\epsilon}$ have zero expected value.
 After removing the terms without contribution to the bias, and to the covariance, with the mathematical expectation operator,
the analytic approximation of the bias (\ref{eqn:biasdef}) is given by 
\begin{equation} \begin{aligned} \mathbf{b}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\mathbf{E}}^2 \left( 2 + 2n - T \right) \mathbf{Q}^{-1} \bm{\theta} , \end{aligned} \label{eqn:biasEu} \end{equation}
and the covariance (\ref{eqn:covdef}) is approximated by 
\begin{equation} \begin{aligned} \mathbf{C}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\bm{\epsilon}}^2 \ \mathbf{Q}^{-1} + \sigma_{\mathbf{E}}^2 \ \mathrm{trace} \left( \bm{\theta} \bm{\theta}^\top \right) \mathbf{Q}^{-1} - \sigma_{\mathbf{E}}^4 \left( 2 + 2n - T \right)^2 \ \mathbf{Q}^{-1} \bm{\theta} \bm{\theta}^\top \mathbf{Q}^{-1} , \end{aligned} \label{eqn:varEu} \end{equation}
where the subscript $\mathrm{p}$ stands for prediction of the bias and covariance.
The derivation of equations (\ref{eqn:biasEu}) and (\ref{eqn:varEu}) is described in Appendix 1.
We use the results described in \citep{Vaccaro94} \S 3 and \citep{Stewart90SPT} \S 2.1 for the expected values of products of unstructured matrices with perturbations.
Equations (\ref{eqn:biasEu}) and (\ref{eqn:varEu}) depend on the unobservable true values $\bm{\theta}$, $\mathbf{K}$, and on the variance of the perturbations.
The observed variables are $\widetilde{\mathbf{y}}$, $\widetilde{\mathbf{K}}$, and from them we compute $\widehat{\bm{\theta}}$.
It is proposed to directly substitute the observed variables in the analytic expressions.
The substitution gives an approximation of the estimation bias and covariance using the observed data.
We have then
\begin{equation} \begin{aligned} \widetilde{\mathbf{b}}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\mathbf{E}}^2 \left( 2 + 2n - T \right) \widetilde{\mathbf{Q}}^{-1} \widehat{\bm{\theta}} , \end{aligned} \label{eqn:biasEuST} \end{equation}
\begin{equation} \begin{aligned} \widetilde{\mathbf{C}}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_\epsilon^2 \ \widetilde{\mathbf{Q}}^{-1} + \sigma_{\mathbf{E}}^2 \ \mathrm{trace} \left( \widehat{\bm{\theta}} \widehat{\bm{\theta}}^\top \right) \widetilde{\mathbf{Q}}^{-1} - \sigma_{\mathbf{E}}^4 \left( 2 + 2n - T \right)^2 \widetilde{\mathbf{Q}}^{-1} \widehat{\bm{\theta}} \widehat{\bm{\theta}}^\top \widetilde{\mathbf{Q}}^{-1} . \end{aligned} \label{eqn:varEuST} \end{equation}
In order to have a prediction of the estimation bias and covariance, the variances $\sigma_{\mathbf{E}}^2$ and $\sigma_{\bm{\epsilon}}^2$ and the observed variables $\widetilde{\mathbf{y}}$, $\widetilde{\mathbf{K}}$, and $\widehat{\bm{\theta}}$ are needed.

@@

The matrices $\mathbf{B}_1$, $\mathbf{B}_2$, $\mathbf{B}_3$, $\mathbf{B}_4$, $\mathbf{C}_1$ and $\mathbf{C}_2$ are considered in the different cases of Lemma \ref{lem:lemma1}.
Each expected value in the lemma is a matrix, or a vector, whose elements are found following the indicated operations.
These operations mainly compute the trace of a product of the corresponding deterministic matrix $\mathbf{A}$, and a matrix constructed from the finite differences matricial operator, which can be of first order $\mathbf{D}^{1}$ or of second order $\mathbf{D}^{2}$.
The total number of operations in the computation of the bias and the covariance is $O\left( T^3 + n^2 \right)$, but this order can be reduced since the $\mathbf{D}$ matrices are sparse.

>>>>

\begin{quote}
%\vspace{-0.5cm}
\emph{The results of this chapter were published in Quintana Carapia, G., Markovsky, I. Pintelon, R., Csurcsia, P.Z., and Verbeke, D., "Bias and covariance of the least-squares estimate in a structured errors-in-variables problem", Computational Statistics \& Data Analysis journal, Vol. 144, 2020, ISSN 0167-9473, doi:10.1016/j.csda.2019.106893. \nocite{QuintanaCSDA} }\vfill{}
\end{quote}

@@

The linear estimation problems with perturbation in both the regression matrix and the regressor are errors-in-variables (EIV) problems.
\color{blue} The formulation of the data-driven step input estimation problem is an EIV problem.
This formulation gives a Hankel structure to the perturbed regression matrix, which is correlated with the regressor vector.
The solution to the structured EIV problem is proposed by means of a recursive least squares algorithm to benefit from its simplicity for online implementations in devices with small computational resources.
The properties of this least-squares (LS) approximation cannot be inferred from classical results\color{black}.
\color{blue} The statistical analysis aims \color{black} to calculate the statistical moments of the \color{blue} least-squares \color{black} estimate to evaluate these properties.
The statistical analysis of the LS estimate demands \color{blue} an additional effort \color{black} when the regression matrix is structured, and when it is correlated with the regressor, in a EIV problem.

\color{blue} 
The total least-squares methods reviewed in \citet{Markovsky07overview} are not appropriate to solve structured EIV problems.
To overcome the challenges imposed by the structure, methods such as the structured total least-squares described in \citet{Lemmerling02} and the structured total maximum likelihood presented in \citet{Beck10} are proposed.
Nevertheless, the structured total least-squares requires optimization and even the recursive alternatives such as the one developed by \citet{Rhode14recursive} are not suitable for online applications in devices with reduced computational resources.
Moreover, current versions of the structured total maximum likelihood method consider uncorrelated cases only. 
The instrumental-variables methods described in \citet{Soderstrom18} are other alternatives to implement recursive algorithms that solve EIV problems.
The generalized instrumental-variables estimator is one example of them, that is mainly used in the identification of linear systems from observed input and output data.
However, it is not straightforward to use the generalized instrumental-variables estimator to implement the direct estimation of the input from the observed output bypassing the model identification, like the data-driven input estimation method does. 
Thus, the recursive LS method is proposed in \citet{Markovsky15cep} to get a real-time approximation to the structured EIV problem solution.
Even though this estimation is suboptimal, we benefit from its simplicity and suitability for implementation in devices with limited computational resources.

As it is described in the works of \citet{Ferrero06} and \citet{Hessling13b}, and in the review of \citet{daSilva12}, the measurement results are meaningful only when the input estimation is provided together with an assessment of its uncertainty.
For an unbiased estimator, the uncertainty is assessed from the estimation variance.
On the other hand, for a biased estimator the mean-squared error is the main metric used to describe the uncertainty.
Therefore, it is necessary to know the first two statistical moments of the estimation.

This chapter describes a methodology to study the statistical moments of the LS solution for the structured EIV problem introduced in \citet{Markovsky15cep, Markovsky15ieee} for the data-driven step input estimation method.
The described statistical analysis obtains the statistical moments of the LS solution, and they allow to assess the estimation uncertainty in terms of the mean-squared error.
The statistical analysis results provide insight into the impact that the structure of the regression matrix and its correlation with the regressor has on the uncertainty of the RLS estimate.

\color{black}

\begin{comment}
Errors-in-variables (EIV) are linear estimation problems in which the regression matrix and the regressor are perturbed \citet{VanHuffel91Book}, \citet{Markovsky07overview}.
In structured EIV problems, the regression matrix has a given structure that depends on the problem formulation.
Hankel, Toeplitz, or other application-specific matrices appear in problems of metrology \citet{Markovsky15cep}, system identification \citet{Soderstrom07}, image restoration \citet{Feiz17}, nuclear magnetic resonance spectroscopy \citet{Cai16}, direction-of-arrival estimation \citet{Pan18}, and time-of-arrival estimation \citet{Jia18}.

In metrology, the direct estimation of the input from the sensor transient response is formulated as a structured EIV problem.
The only observed signal is the sensor output.
To estimate a step input, the regression matrix, and the regressor are built from the step response observations \citet{Markovsky15cep}, and
the structure in the regression matrix is block-Hankel.
This data-driven estimation methodology reduces the estimation time of the classical two-stage approach where a sensor model is first identified, and later the input is estimated using the sensor model \citet{Azam15, Niedzwiecki16a}.
%The problem is EIV because the perturbation of the sensor response gets into the regression matrix.

Total least-squares (TLS) is the typical estimator for unstructured EIV problems and is consistent when the perturbations have zero mean, and the covariance is a given positive definite matrix.
When the perturbations are i.i.d. normally distributed, the solution of the TLS is equivalent to that of the maximum likelihood estimator (ML) \citet{Markovsky07overview}. 
For structured EIV problems, the TLS estimator does not give general results since each specific structure requires a particular treatment \citet{VanHuffel07TLSeditorial}, and the ML estimator leads to non-convex optimization problems where finding the global optimum is not guaranteed \citet{Rhode14recursive}.
Moreover, the computational complexity of TLS and ML inhibits real-time implementation to solve structured EIV problems.

The least-squares (LS) estimator is a suboptimal but simple solution to structured EIV problems that admits a recursive form for easy real-time implementations.
Some of the reported works that propose LS estimators for structured EIV problems include 
the design of a fast algorithm for matrices with small displacement rank \citet{Mastronardi07fast},
the study of the estimator consistency \citet{Palanthandalam10parameter},
the determination of the bias, and the mean squared error of the parameter estimates in the identification of AR models \citet{Kiviet12high} \citet{Kiviet14improved}, and
a discussion of the causes of bias and inconsistency in homogeneous estimators \citet{Yeredor04homogeneous}.

 In measurement applications, it is highly relevant to assess the uncertainty of the input estimate.
The uncertainty of the reported LS estimators for structured EIV problems has not been addressed, and then, remains unknown.
The estimator uncertainty is expressed using the estimation bias and covariance \citet{Pintelon12Book}.
To know the LS estimator uncertainty, we quantified the bias and the covariance of the LS solution of EIV problems in the unstructured and structured cases. 
This extends the perturbation analysis of the LS estimator of unstructured and uncorrelated problems that was investigated in \citet{Stewart90SPT} and in \citet{Vaccaro94}.
This paper presents a study of the statistics of the LS estimate of unstructured and structured EIV problems. 
 We provide a discussion of the unstructured case as a reference, to get an insight of the impact that the structure of the regression matrix and the correlation between the regression matrix and the regressor perturbations have on the uncertainty of the LS estimate.
The structured case is motivated by the metrology estimation problem \citet{Markovsky15cep}, where the EIV problem has a Hankel structure, and the perturbations are correlated.
The study of the metrology input estimate illustrates a methodology to conduct statistical analysis for any structured EIV problem.
The mathematical expectation of the second-order Taylor series expansion of the LS estimate boils down to expressions that quantify the first and second-order moments of the LS estimate.
Via Monte Carlo simulations, we validated the accuracy of the bias, and the covariance approximations.
The derived approximations predict the LS estimation bias, and the covariance, for given sample size and perturbation level.
The predicted variance gives the uncertainty of the LS estimate.
We observed that, for the step input estimation problem, the mean squared error of the LS estimate is near to the minimum variance limit given by the Cram\'er-Rao lower bound of the structured EIV problem.
 By following this methodology, the bias and variance of the solutions of EIV problems with other structures is determined, and therefore, the uncertainty of the estimate.
\end{comment}

@@

For an overdetermined system of linear equations, constructed by the data-driven step input estimation method (\ref{eqn:min_seiv}), the least-squares (LS) solution is given by
\ref{eqn:xhat}.
The objective of the statistical analysis is to obtain the bias and the covariance of the solution $\widehat{\bm{\theta}}$ to the structured and correlated EIV problem (\ref{eqn:ddsiemnd}).
The bias and the covariance are computed using the mathematical expectation operator $\mathbb{E}\{\cdot\}$.
First we substitute the expressions (\ref{eqn:y0plusnoise}) and (\ref{eqn:K0plusnoise}) in Equation (\ref{eqn:xhat}) to obtain
\begin{equation*} \widehat{\bm{\theta}} = \left( (\mathbf{K}+\mathbf{E})^\top (\mathbf{K}+\mathbf{E})  \right)^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}), \end{equation*} 
which is equivalent to \color{blue}
\begin{equation*} \widehat{\bm{\theta}} = \left( \left( \mathbf{K}^\top \mathbf{K} \right)  \left( \mathbf{I} + \left( \mathbf{K}^\top \mathbf{K} \right)^{-1} \left( \mathbf{K}^\top \mathbf{E} + \mathbf{E}^\top \mathbf{K} + \mathbf{E}^\top \mathbf{E} \right) \right)  \right)^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}), \end{equation*} 
and \color{black}
\begin{equation} \widehat{\bm{\theta}} = \left( \mathbf{I} + \mathbf{M} \right)^{-1} \mathbf{Q}^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}), \label{eqn:1steq} \end{equation} 
where 
\begin{equation} \mathbf{Q} = \mathbf{K}^\top \mathbf{K}, \quad \text{and} \quad \mathbf{M} = \mathbf{Q}^{-1} ( \mathbf{K}^\top \mathbf{E} + \mathbf{E}^\top \mathbf{K} + \mathbf{E}^\top \mathbf{E} ). \end{equation} 
Applying a second-order Taylor expansion of the inverse matrix
\begin{equation} (\mathbf{I} + \mathbf{M})^{-1} \approx \mathbf{I} - \mathbf{M} + \mathbf{M}^2, \label{eqn:TseriesExp} \end{equation} 
that is valid when the SNR is sufficiently high, and therefore $\mathbf{E}$ and $\mathbf{M}$ are small, satisfying the constraint on the spectral radius $\color{blue} \rho(\mathbf{M}) = \color{black} \| \mathbf{M} \| < 1$. 
The neglected term in the Taylor series expansion is of the order $O(\Vert M \Vert^3)$.
\color{blue} An a priori bound on the error introduced by the second order Taylor series expansion can be expressed in terms of the spectral radius $\rho(\mathbf{M})$.
Considering that the Taylor series expansion is an infinite sum we have that
\begin{equation} \begin{aligned} (\mathbf{I} + \mathbf{M})^{-1} &= \mathbf{I} - \mathbf{M} + \mathbf{M}^2 - \mathbf{M}^3 + \ldots \\
&=  \mathbf{I} - \mathbf{M} + \mathbf{M}^2 - \mathbf{M}^3 \left( \mathbf{I} - \mathbf{M} + \mathbf{M}^2 - \mathbf{M}^3 + \ldots \right) \\
&=  \mathbf{I} - \mathbf{M} + \mathbf{M}^2 - \mathbf{M}^3  (\mathbf{I} + \mathbf{M})^{-1} \end{aligned} \end{equation} 
Therefore we can obtain a bound for the second error expansion error if we take a matrix norm as follows
\begin{equation} \begin{aligned} \| (\mathbf{I} + \mathbf{M})^{-1} - ( \mathbf{I} - \mathbf{M} + \mathbf{M}^2 ) \| &=  \| \mathbf{M}^3  (\mathbf{I} + \mathbf{M})^{-1} \| \\
& \leq \| \mathbf{M}^3 \| \| (\mathbf{I} + \mathbf{M})^{-1} \| \leq \dfrac{\| \mathbf{M} \|^3}{ 1 -  \| \mathbf{M} \|} = \dfrac{\rho(\mathbf{M})^3}{1 - \rho(\mathbf{M})}  \end{aligned} \end{equation}

Using the second order Taylor expansion of the inverse matrix\color{black}, we can express the estimate as
\begin{equation} \widehat{\bm{\theta}} \approx \left( \mathbf{I} - \mathbf{M} + \mathbf{M}^2 \right) \mathbf{Q}^{-1} (\mathbf{K}+\mathbf{E})^\top (\mathbf{y}+\bm{\epsilon}). \label{eqn:xhatexp} \end{equation}

Now that the perturbation variables $\bm{\epsilon}$ and $\mathbf{E}$ are no longer inside an inverse matrix, we can compute the mathematical expectation of expressions derived from the Taylor series approximation (\ref{eqn:xhatexp}) of $\widehat{\bm{\theta}}$. 
The bias of the estimate $\widehat{\bm{\theta}}$ is obtained from
\begin{equation} \mathbf{b} \left(\widehat{\bm{\theta}} \right) = \bm{\mu}\left(\widehat{\bm{\theta}} \right) - \bm{\theta}, \label{eqn:biasdef} \end{equation}
where $\bm{\mu} \left(\widehat{\bm{\theta}} \right) = \mathbb{E} \left\{ \widehat{\bm{\theta}} \right\}$, and $\bm{\theta} = \mathbf{K}^\dagger \mathbf{y} = \mathbf{Q}^{-1} \mathbf{K}^\top \mathbf{y}$ is the true value.
The covariance of the estimate $\widehat{\bm{\theta}}$ is obtained from
\begin{equation} \begin{aligned} \mathbf{C} \left( \widehat{\bm{\theta}} \right) & = \mathbb{E} \left\{ \left( \widehat{\bm{\theta}} - \bm{\mu} \color{blue} \left(\widehat{\bm{\theta}} \right) \color{black} \right) \left( \widehat{\bm{\theta}} - \bm{\mu} \color{blue} \left(\widehat{\bm{\theta}} \right) \color{black} \right)^\top \right\} . \end{aligned} \label{eqn:covdef} \end{equation}

The terms derived from (\ref{eqn:xhatexp}) that do not contribute to the bias and to the covariance are filtered out by the mathematical expectation operator, described in \citet{Papoulis02Book}, considering the following general rules that are valid regardless of the existence of structure in the regression problem: 
\begin{itemize}
	\item the expected values $\mathbb{E} \{ \mathbf{E} \} = \mathbf{0}$, and $\mathbb{E} \{ \bm{\epsilon} \} = \mathbf{0}$, since $\mathbf{E}$ and $\bm{\epsilon}$ are zero-mean random variables, and
	\item the expected values of odd order moments, such as $\mathbb{E} \{ \mathbf{E}^\top \mathbf{E} \mathbf{E}^\top \}$, are zero.
\end{itemize}
Moreover, the second-order approximation disregards moments of order four and higher.

\color{blue}The normality assumption of the perturbation noise is necessary to provide more prior information into the method formulation.
The second order Taylor series expansion (\ref{eqn:xhatexp}) is developed considering the perturbation noise that gets into the regression matrix from the sensor transient response.
Assuming only that the perturbation noise is distributed with zero mean and variance $\sigma_{\bm{\epsilon}}^2$ is not enough.
If we additionally assume normality for the perturbation noise, then we benefit from the knowledge of the third moment equal to zero due to symmetry, and the fourth moment equal to three times the squared variance\color{black}.

@@

First, the statistics of the least-squares estimator (\ref{eqn:xhat}) of an unstructured EIV problem is discussed, 
provided that the perturbations of the regression matrix and the regressor are i.i.d. normally distributed with zero mean, and variances $\sigma_{\mathbf{E}}^2$ and $\sigma_{\bm{\epsilon}}^2$, respectively.
Therefore, the terms in the Taylor series expansion that contain products of $\mathbf{E}$ and $\bm{\epsilon}$ have zero expected value.
After removing the terms without contribution to the bias, and to the covariance, with the mathematical expectation operator,
the analytic approximation of the bias (\ref{eqn:biasdef}) is given by 
\begin{equation} \begin{aligned} \mathbf{b}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\mathbf{E}}^2 \left( 2 + 2n - \color{blue}N\color{black} \right) \mathbf{Q}^{-1} \bm{\theta} , \end{aligned} \label{eqn:biasEu} \end{equation}
and the covariance (\ref{eqn:covdef}) is approximated by 
\begin{equation} \begin{aligned} \mathbf{C}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\bm{\epsilon}}^2 \ \mathbf{Q}^{-1} + \sigma_{\mathbf{E}}^2 \ \mathrm{trace} \left( \bm{\theta} \bm{\theta}^\top \right) \mathbf{Q}^{-1} - \sigma_{\mathbf{E}}^4 \left( 2 + 2n - \color{blue}N\color{black} \right)^2 \ \mathbf{Q}^{-1} \bm{\theta} \bm{\theta}^\top \mathbf{Q}^{-1} , \end{aligned} \label{eqn:varEu} \end{equation}
where the subscript $\mathrm{p}$ stands for prediction of the bias and covariance.
The derivation of equations (\ref{eqn:biasEu}) and (\ref{eqn:varEu}) is described in Appendix 1.
We use the results described in \citet{Vaccaro94} \S 3 and \citet{Stewart90SPT} \S 2.1 for the expected values of products of unstructured matrices with perturbations.
Equations (\ref{eqn:biasEu}) and (\ref{eqn:varEu}) depend on the unobservable true values $\bm{\theta}$, $\mathbf{K}$, and on the variance of the perturbations.
The observed variables are $\widetilde{\mathbf{y}}$, $\widetilde{\mathbf{K}}$, and from them we compute $\widehat{\bm{\theta}}$.
It is proposed to directly substitute the observed variables in the analytic expressions.
The substitution gives an approximation of the estimation bias and covariance using the observed data.
We have then
\begin{equation} \begin{aligned} \widetilde{\mathbf{b}}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_{\mathbf{E}}^2 \left( 2 + 2n - \color{blue}N\color{black} \right) \widetilde{\mathbf{Q}}^{-1} \widehat{\bm{\theta}} , \end{aligned} \label{eqn:biasEuST} \end{equation}
\begin{equation} \begin{aligned} \widetilde{\mathbf{C}}_{\mathrm{p}} \left( \widehat{\bm{\theta}} \right) & \approx \sigma_\epsilon^2 \ \widetilde{\mathbf{Q}}^{-1} + \sigma_{\mathbf{E}}^2 \ \mathrm{trace} \left( \widehat{\bm{\theta}} \widehat{\bm{\theta}}^\top \right) \widetilde{\mathbf{Q}}^{-1} - \sigma_{\mathbf{E}}^4 \left( 2 + 2n - \color{blue}N\color{black} \right)^2 \widetilde{\mathbf{Q}}^{-1} \widehat{\bm{\theta}} \widehat{\bm{\theta}}^\top \widetilde{\mathbf{Q}}^{-1} . \end{aligned} \label{eqn:varEuST} \end{equation}
In order to have a prediction of the estimation bias and covariance, the variances $\sigma_{\mathbf{E}}^2$ and $\sigma_{\bm{\epsilon}}^2$ and the observed variables $\widetilde{\mathbf{y}}$, $\widetilde{\mathbf{K}}$, and $\widehat{\bm{\theta}}$ are needed.

@@

The matrices $\mathbf{B}_1$, $\mathbf{B}_2$, $\mathbf{B}_3$, $\mathbf{B}_4$, $\mathbf{C}_1$ and $\mathbf{C}_2$ are considered in the different cases of Lemma \ref{lem:lemma1}.
Each expected value in the lemma is a matrix, or a vector, whose elements are found following the indicated operations.
These operations mainly compute the trace of a product of the corresponding deterministic matrix $\mathbf{A}$, and a matrix constructed from the finite differences matricial operator, which can be of first order $\mathbf{D}^{1}$ or of second order $\mathbf{D}^{2}$.
The total number of operations in the computation of the bias and the covariance is $O\left( \color{blue}N\color{black}^3 + n^2 \right)$, but this order can be reduced since the $\mathbf{D}$ matrices are sparse.
